Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-13 surfaces-write-up/surface-write-up.tex
edits to section 5 and 4?

\DeclareMathOperator\di{Div}
\newcommand\bida{a}
\newcommand\bidb{b}
\newcommand\pdeg{\delta}
\newcommand\hirz{\mathbb{F}}
\newcommand\mss{\mathscr{S}}

@@

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D, x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Then, define
\begin{align*}
	\Sigma = \langle u^d y_1^{c_1} \cdots y_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{} \bidb_i c_i. \rangle 
\end{align*}
In this case, we have a factorization of $\phi$ given by maps $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
$$\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}y_1^{c_{i1}} \cdots y_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}$$
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
$$\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0
\end{tikzcd}.$$
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

@@

\begin{defn}
\label{defn:vec-sum}
For $\vec{v} \in \br^n$, denote $\deg \vec{v} := \sum_{i = 0}
^n v_i$ (i.e. the taxicab metric).
Let $\mss_i := \{\vec{v} \in \bz_{\geq 0}^{m + 1} \; : \;
\sum_{i = 0} ^{m + 1} v_i = c_i\}$ be the set of $\bz_{\geq 0}
^{m + 1}$-vectors with sum of components equal to $c_i$.
\end{defn}

\begin{defn}
\label{defn:vec-order}
Let $\vec{v}, \vec{w} \in \br^n$ be two vectors. Let $i$
be the right-most coordinate such that $v_i$ is nonzero
and $j$ be the left-most coordinate such that $w_j$ is
nonzero. Define $\vec{v} \prec \vec{w}$ if $i \leq j$.
\end{defn}

\begin{thm}
\label{thm:proj-one-point}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \alpha H$, with $\alpha = \frac{p}{q} \in \bq$
and $H = V(x_k)$ a hyperplane of $\bp^m$, be a divisor on $\bp^m$.
Let
\[
	0 = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

@@

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + i} \choose {i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $\vec{v} \in \bz_{\geq 0}^{m + 1}$
such that $\sum_{j = 0}^{m} v_j = c_i$. Furthermore, it has $z$
many \todo{how many?} relations of the following two types.

For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{k_{i, j}}} (F_{k_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < k_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\substack{\vec{y} \in \mss_{k_{i, j}} \\ \vec{z} \in
	\mss_{k_{i, j} + 1}}} (F_{k_{i, j}}^{\vec{y}})^{g_{\vec{y}}} \;
	(F_{k_{i, j} + 1}^{\vec{z}})^{h_{\vec{z}}}
	&(i < k_{i, j} < k_{i, j} + 1 < j).
\end{flalign*}

Furthermore, there are the relations for each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) of the form 
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{w}} \\
\end{align*}

\noindent
where $\vec{y} \in \mss_i$ and $\vec{w} \in \mss_j$ are the unique
vectors such that $\vec{y} \prec \vec{w}$.
\end{thm}

\begin{proof}
We know $\left\{F_{0}^{\vec{v}} : \deg \vec{v} = \left\lfloor
\frac{p}{q} \right\rfloor \right\}$ are minimal generators in
degree $1$. Now let $i > 0$. Suppose $F_i^{\vec{v}}$ were not
minimal for some $\vec{v} \in \mss_i$. Then $F_i^{\vec{v}} =
F_{s}^{\vec{w}} F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and
$\vec{w}, \vec{z} \in \bz_{ \geq 0}^{m + 1}$. In particular, we see
that 
\begin{align*}
	&d_i = s + t \\
	&c_i = \deg \vec{v} = \deg \vec{w} + \deg \vec{z}
\end{align*}

@@

Now we show that no further generators are necessary. As done in
the proof of the $\bp^1$ case by O'Dorney \cite[Theorem 6]
{dorney:canonical}, each lattice point $(\beta, \gamma) \in
\bz_{\geq 0}^2$ such that $c \leq d \alpha$ is a non-negative
linear combination of $\vec{x}_i$ and $\vec{x}_{i + 1}$ where
$\vec{x}_i := (d_i, c_i) \in \bz^2$. Consequently, if $(\beta,
\gamma) = \ell_1 (d_i, c_i) + \ell_2 (d_{i + 1}, c_{i + 1})$ for
$\ell_1, \ell_2 \in \bz_{ \geq 0}$, then any element $\frac{u^{\beta}
x^{\vec{v}}} {x_k^{ \gamma}} \in R_D$ is expressible as
\begin{align*}
	\frac{u^{\beta} x^{\vec{v}}} {x_k^{\gamma}} = \left(\frac{u^{d_i}
	x^{\vec{w}}} {x_k^{c_i}}\right)^{\ell_1} \left(\frac{u^{d_{i + 1}}
	x^{\vec{z}}} {x_k^{c_{i + 1}}}\right)^{\ell_2}
\end{align*}

\noindent
for some $\vec{w} \in \mss_i, \vec{z} \in \mss_{i + 1}$. But these
are precisely powers of two generators of the form $F_i^{\vec{w}}$
and $F_{d_{i + 1}}^{\vec{z}}$ as prescribed in the theorem statement.

@@

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \todo{cite earlier 
proposition}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sigma = \sum_{i\in T_=(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{i\in T_+(D) \atop
	j \in T_-(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

>>>>

\DeclareMathOperator\di{Div}
\newcommand\bida{a}
\newcommand\bidb{b}
\newcommand\pdeg{\delta}
\newcommand\hirz{\mathbb{F}}

@@

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D$ given by $x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Then, define
\begin{align*}
	\Sigma = \langle u^d y_1^{c_1} \cdots y_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i. \rangle 
\end{align*}
In this case, we can factor $\phi$ through map;s $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
$$\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}y_1^{c_{i1}} \cdots y_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}$$
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
$$\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0
\end{tikzcd}.$$
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

@@

\begin{thm}
\label{thm:proj-one-point}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \alpha H$, with $\alpha = \frac{p}{q} \in \bq$
and $H = V(x_k)$ a hyperplane of $\bp^m$, be a divisor on $\bp^m$.
Let
\[
	\left\lfloor \frac{p}{q} \right\rfloor = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

@@

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + i} \choose {i}}$ generators $F_{\vec{v}}^{d_i} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $\vec{v} \in \bz_{\geq 0}^{m + 1}$
such that $\sum_{j = 0}^{m} v_j = c_i$ and $z$ \todo{how many?}
relations of the form
\[
	G_{d_i, d_j}^{\vec{v}, \vec{w}} = F_{d_i}^{\vec{v}} F_{d_j}^{\vec{w}}
	- (F_{d_k}^{\vec{y}})^{g_{i, j}} (F_{d_\ell}^{\vec{z}})^{g_{i, j}}
\]

\noindent
when $d_i + d_j = d_k + d_\ell$ and $\vec{v} + \vec{w} = \vec{y} +
\vec{z}$.
\end{thm}

\begin{proof}
We know $\left\{F_{0}^{\vec{v}} : \deg \vec{v} = \left\lfloor \frac{p}{q}
\right\rfloor \right\}$ are minimal generators in degree $1$. Now let $i >
0$. If $F_{d_i}^{\vec{v}}$ were not minimal for some $\vec{v}$ such
that $\deg \vec{v} = c_i$, then $F_{d_i}^{\vec{v}} = F_{s}^{\vec{w}}
F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and $\vec{w}, \vec{z} \in
\bz_{\geq 0}^{m + 1}$. In particular, we see that 
\begin{align*}
	&d_i = s + t \\
	&c_i = \deg \vec{v} = \deg \vec{w} + \deg \vec{z}
\end{align*}

@@

Now we show that no further generators are necessary. As done in
the proof of the $\bp^1$ case by O'Dorney \cite[Theorem 6]
{dorney:canonical}, each lattice point $(\beta, \gamma) \in
\bz_{\geq 0}^2$ such that $c \leq d \alpha$ is a non-negative
linear combination of $\vec{x}_i$ and $\vec{x}_{i + 1}$ where
$\vec{x}_i := (d_i, c_i) \in \bz^2$. Consequently, if $(\beta,
\gamma) = h _1 (d_i, c_i) + h_2 (d_{i + 1}, c_{i + 1})$ for $h_1,
h_2 \in \bz_{ \geq 0}$, then any element $\frac{u^{\beta} x^{\vec{v}}}
{x_k^{ \gamma}} \in R_D$ is expressible as
\begin{align*}
	\frac{u^{\beta} x^{\vec{v}}} {x_k^{\gamma}} = (\frac{u^{d_i} x^{\vec{w}}}
	{x_k^{c_i}})^{h_1} (\frac{u^{d_{i + 1}} x^{\vec{z}}} {x_k^{c_{i + 1}}})^{h_2}
\end{align*}

\noindent
for some $\vec{w}, \vec{z} \in \bz_{\geq 0}^{m + 1}$ such that $\deg
\vec{w} = c_i, \deg \vec{z} = c_{i + 1}$. But these are precisely
powers of two generators of the form $F_{d_i}^{\vec{w}}$ and
$F_{d_{i + 1}}^{\vec{z}}$ as prescribed in the theorem statement.

@@

\begin{lem}
Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^n$.  Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$.  
\end{lem}

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \ref{prop:cone-generation}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sigma = \sum_{i\in T_=(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{i\in T_+(D) \atop
	j \in T_-(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}
